\cursedforest\ - A Random Forest Implementation for ``Big'' and ``Wide'' Data \\ Supplementary Information

### 2017-02-21 doc/publications/plos16/plos16.tex
changes to conclusion

We have demonstrated that using a different parallelization model can extend random forests to the case of an
extremely large number of variables. We have treated the case of variable selection in a $p >> n$ model where most of
the variables are uninformative and have demonstrated the utility of the model for large GWAS datasets. By comparing
this implementation to other implementations (including those optimized for large datasets) we have demonstrated the
utility of this approach.

Donoho and Tanner \cite{Donoho.and.Tanner.2009} give a ``universal phase change'' result that has applications in a
large number of areas including variable selection in high dimensions. They show that there is a sharp disjunction
between the cases where significant variables may be recovered with a high accuracy by procedures like stepwise variable
selection, and the cases where they can not be recovered. The boundary is shown in
Fig~\ref{figure:phase-diagram-equivalence.png} in a space parameterized by the level of underdetermination,
$\delta = n/p$, and by the sparsity, $\rho =k/n$ (where $k$ is the number of significant variables).  Above the
phase-transition line variable recovery is still possible by a combinatorial search approach such as all-subsets
variable selection.  The theoretical boundary is based on arguments from combinatorial geometry.

@@

\cite{Donoho.and.Stodden.2006} investigate the phase transition in a small simulation.  They consider a regression problem with 
$\underset{n\times p}{X}\sim N(0,1)$ with $p$ fixed at 200 and $n$ variable.
They  set $\beta(1:k) \sim U(1,100)$ and $\beta((k+1):p) =0$, then  $y= X\beta + \epsilon$, $\epsilon \sim
N(0,\sqrt{16})$. They evaluate variable selection by the error measure 
$$\frac{\Vert\hat{\beta}-\beta\Vert_2}{\Vert\beta\Vert_2}.$$

@@

 \begin{figure}[tbhp] 
%   \begin{adjustwidth}{-1.00in}{0in}
     \begin{subfigure}[t]{0.5\linewidth}
       \centering
       \includegraphics[totalheight=6cm]{./figs/error_Stodden_FDR.png}
       \caption{The error measure.}
       \label{figure:error_Stodden_FDR.png}
%       \vspace{4ex}
     \end{subfigure} 
     \begin{subfigure}[t]{0.5\linewidth}
       \centering
       \includegraphics[totalheight=6cm]{./figs/rbo_Stodden_FDR.png}
       \caption{The RBO measure}
       \label{figure:rbo_Stodden_FDR.png}
     \end{subfigure} 
     \caption{Forward stepwise variable selection with a false discovery rate stopping criteria.}
       \label{figure:error_Stodden_FDR.png}
%   \end{adjustwidth}
 \end{figure}

 \begin{figure}[tbhp] 
%   \begin{adjustwidth}{-1.00in}{0in}
     \begin{subfigure}[t]{0.5\linewidth}
       \centering
       \includegraphics[totalheight=6cm]{./figs/ranger_error_Stodden_simulation.png}
       \caption{The OOB error rate.}
       \label{figure:ranger_error_Stodden_simulation.png}
       \vspace{4ex}
     \end{subfigure} 
     \begin{subfigure}[t]{0.5\linewidth}
       \centering
       \includegraphics[totalheight=6cm]{./figs/ranger_rbo_Stodden_simulation.png}
       \caption{The RBO measure}
       \label{figure:ranger_rbo_Stodden_simulation.png}
     \end{subfigure} 
     \caption{The simulated data of \protect{\cite{Donoho.and.Stodden.2006}} with a random forest model.}
     \label{figure:ranger_rbo_Stodden_simulation.png}
%   \end{adjustwidth}
 \end{figure}

@@

We have used a random forest on the same simulation as figure \ref{figure:error_Stodden_FDR.png}.  The Donoho-Tanner phase transition arises in
recovering the $\beta$ in data generated by a linear model. However, in a decision tree (random forest) we are fitting a
non-linear model and there is no notion of estimating the $\beta$. Because of this we have evaluated the performance of
the random forest using the RBO measure on the variable importance.

@@

Fig~\ref{figure:ranger_rbo_Stodden_simulation.png} shows the OOB prediction error and the RBO error for a random forest with 10000 trees and the \mtry value set
to the default for a regression problem (depending on $n$). It shows no evidence of following the shape of the
Donoho-Tanner phase transition.

As \cursedforest is designed for extremely large numbers of variables it is likely to be operating in difficult regions
of the figure where the ratio $\delta = n/p$ is small. In the simulated example in section \ref{section:synthetic_data}
the underdetermination $\delta = 0.002$ and the sparsity $\rho =2\times 10^{-6}.$ The existence of the Donoho-Tanner
phase transition is a salutary warning. There are likely to be limits, both computational and logical, to the recovery
of signals from noisy data. \cursedforest is a contribution to addressing the practical limits but the logical limits
will still apply. However in the case of data that is both big and wide, \cursedforest and other VariantSpark methods
may provide a useful tool.

>>>>

We have demonstrated that using a different parallelization model can extend random forests to the case of an
extremely large number of variables. We have treated the case of variable selection in a $p >> n$ model, where most of
the variables are uninformative, and have demonstrated the utility of the model for large GWAS datasets. By comparing
this implementation to other implementations (including those optimized for large datasets) we have demonstrated the
utility of this approach.

Donoho and Tanner \cite{Donoho.and.Tanner.2009} give a ``universal phase change'' result that has applications in a
large number of areas, including variable selection in high dimensions. The theoretical phase change boundary is based on arguments
from combinatorial geometry.

They argue that there is a sharp disjunction between the cases where informative variables may be recovered with a high
accuracy by procedures like stepwise variable selection, and the cases where they can not be recovered. The boundary is
shown in Fig~\ref{figure:phase-diagram-equivalence.png} in a space parameterized by the level of underdetermination,
$\delta = n/p$, and by the sparsity, $\rho =k/n$ (where $k$ is the number of informative variables).  Above the
phase-transition line variable recovery is still possible by a combinatorial search approach such as all-subsets
variable selection.

@@

\cite{Donoho.and.Stodden.2006} investigate the phase transition in a small simulation.  They consider a regression
problem with $\underset{n\times p}{X}\sim N(0,1)$ with $p$ fixed at 200 and $n$ variable.  They set
$\beta(1:k) \sim U(1,100)$ and $\beta((k+1):p) =0$, then $y= X\beta + \epsilon$ with $\epsilon \sim N(0,\sqrt{16})$. They
evaluate variable selection by the error measure
$$\frac{\Vert\hat{\beta}-\beta\Vert_2}{\Vert\beta\Vert_2}.$$

@@

 \begin{figure}[tbhp] 
%   \begin{adjustwidth}{-1.00in}{0in}
     \begin{subfigure}[t]{0.5\linewidth}
       \centering
       \includegraphics[totalheight=6cm]{./figs/error_Stodden_FDR.png}
       \caption{The error measure.}
       \label{figure:error_Stodden_FDR.png}
%       \vspace{4ex}
     \end{subfigure} 
     \begin{subfigure}[t]{0.5\linewidth}
       \centering
       \includegraphics[totalheight=6cm]{./figs/rbo_Stodden_FDR.png}
       \caption{The RBO measure}
       \label{figure:rbo_Stodden_FDR.png}
     \end{subfigure} 
     \caption{Forward stepwise variable selection with a false discovery rate stopping criteria.}
       \label{figure:error_and_rbo_Stodden_FDR.png}
%   \end{adjustwidth}
 \end{figure}

 \begin{figure}[tbhp] 
%   \begin{adjustwidth}{-1.00in}{0in}
     \begin{subfigure}[t]{0.5\linewidth}
       \centering
       \includegraphics[totalheight=6cm]{./figs/ranger_error_Stodden_simulation.png}
       \caption{The OOB error rate.}
       \label{figure:ranger_error_Stodden_simulation.png}
       \vspace{4ex}
     \end{subfigure} 
     \begin{subfigure}[t]{0.5\linewidth}
       \centering
       \includegraphics[totalheight=6cm]{./figs/ranger_rbo_Stodden_simulation.png}
       \caption{The RBO measure}
       \label{figure:ranger_rbo_Stodden_simulation.png}
     \end{subfigure} 
     \caption{The simulated data of \protect{\cite{Donoho.and.Stodden.2006}} with a random forest model.}
     \label{figure:ranger_error_and_rbo_Stodden_simulation.png}
%   \end{adjustwidth}
 \end{figure}

@@

We have used a random forest on the same simulation data used for figure \ref{figure:error_and_rbo_Stodden_FDR.png}.  The
Donoho-Tanner phase transition arises in recovering the $\beta$ in data generated by a linear model. However, in a
decision tree (random forest) we are fitting a non-linear model and there is no notion of estimating the
$\beta$. Because of this we have evaluated the performance of the random forest using the RBO measure on the variable
importance.

@@

Fig~\ref{figure:ranger_error_and_rbo_Stodden_simulation.png} shows the OOB prediction error and the RBO error for a
random forest with 10000 trees and the \mtry value set to the default for a regression problem (depending on $n$). It
shows no evidence of following the shape of the Donoho-Tanner phase transition.

As \cursedforest is designed for extremely large numbers of variables it is likely to be operating in difficult regions
of the figure where the ratio $\delta = n/p$ is small. In the simulated example in section \ref{section:synthetic_data}
the underdetermination parameter $\delta = 0.002$ and the sparsity parameter $\rho =2\times 10^{-6}.$

The existence of the Donoho-Tanner phase transition is a salutary warning. There are likely to be limits, both
computational and logical, to the recovery of signals from noisy data. \cursedforest is a contribution to addressing the
practical limits but the logical limits will still apply. However in the case of data that is both big and wide,
\cursedforest and other VariantSpark methods may provide a useful tool.
